---
title: "GUT MICROBIOME ML PROJECT parkinson"
output: html_document
date: "2026-04-12"
---

otu_table()Count matrixHow many of each bacteria per sample

sample_data()MetadataInfo ABOUT each patient

tax_table()TaxonomyWhat each bacteria is called

phy_tree()TreeEvolutionary relationships



in terminal 

ls Wallen_DAMethodCompare_2021/Dataset1_Scripts/



library(phyloseq)

Dataset1 <- readRDS("Wallen_DAMethodCompare_2021/PhyloseqObjects/Dataset1/phyloseq.rds")

Dataset2 <- readRDS("Wallen_DAMethodCompare_2021/PhyloseqObjects/Dataset2/phyloseq.rds")

```{r setup,cache=TRUE}
# Step 1: Explore 
#FOR SAMPLE_DATA


table(sample_data(Dataset1)$case_control)

colnames(sample_data(Dataset1))

head(sample_data(Dataset1))


```
Sample Data:        [6 samples by 3 sample variables]:
                NGRC_SampleID        SampleID case_control
10122.EC0013201     EC0013201 10122.EC0013201      Control
10122.EC0037201     EC0037201 10122.EC0037201      Control
10122.EC0044201     EC0044201 10122.EC0044201      Control
10122.EC0046201     EC0046201 10122.EC0046201      Control
10122.EC0051201     EC0051201 10122.EC0051201      Control


```{r setup,cache=TRUE}
#EXPLORE FOR OTU TABLE
otu_table(Dataset1)[1:5, 1:5]

# Total reads per sample
head(colSums(otu_table(Dataset1)))

# Total reads per taxon
head(rowSums(otu_table(Dataset1)))

# Dimensions
dim(otu_table(Dataset1))

```


```{r setup,cache=TRUE}
#EXPLORE

head(tax_table(Dataset1))


```



```{r setup,cache=TRUE}

library(ggplot2)


reads_df_1<-data.frame(
  reads=sample_sums(Dataset1),
  group=sample_data(Dataset1)$case_control
)

p1<-ggplot(reads_df, aes(x = reads, fill = group)) +
  geom_histogram(bins = 30, alpha = 0.7, color = "white") +
  scale_fill_manual(values = c("Case" = "#E74C3C", "Control" = "#3498DB")) +
  theme_minimal(base_size = 13) +
  labs(title = "Read Depth per Sample",
       x = "Total Reads", y = "Number of Samples", fill = "Group")

print(p)


#FOR DATASET 2 



reads_df_2<-data.frame(
  reads=sample_sums(Dataset2),
  group=sample_data(Dataset2)$case_control
)

p2<-ggplot(reads_df_2, aes(x = reads, fill = group)) +
  geom_histogram(bins = 30, alpha = 0.7, color = "white") +
  scale_fill_manual(values = c("Case" = "#E74C3C", "Control" = "#3498DB")) +
  theme_minimal(base_size = 13) +
  labs(title = "Read Depth per Sample",
       x = "Total Reads", y = "Number of Samples", fill = "Group")

#side by side
library(gridExtra)
merged<-grid.arrange(p1,p2,ncol=2)

print(merged)

ggsave("read_depth_both.png", plot = merged, width = 12, height = 5)





```