---
title: "GUT MICROBIOME ML PROJECT parkinson"
output: html_document
date: "2026-04-12"
---

otu_table()Count matrixHow many of each bacteria per sample

sample_data()MetadataInfo ABOUT each patient

tax_table()TaxonomyWhat each bacteria is called

phy_tree()TreeEvolutionary relationships



in terminal 

ls Wallen_DAMethodCompare_2021/Dataset1_Scripts/



library(phyloseq)

Dataset1 <- readRDS("Wallen_DAMethodCompare_2021/PhyloseqObjects/Dataset1/phyloseq.rds")

Dataset2 <- readRDS("Wallen_DAMethodCompare_2021/PhyloseqObjects/Dataset2/phyloseq.rds")

```{r setup,cache=TRUE}
# Step 1: Explore 
#FOR SAMPLE_DATA


table(sample_data(Dataset1)$case_control)

colnames(sample_data(Dataset1))

head(sample_data(Dataset1))


```
Sample Data:        [6 samples by 3 sample variables]:
                NGRC_SampleID        SampleID case_control
10122.EC0013201     EC0013201 10122.EC0013201      Control
10122.EC0037201     EC0037201 10122.EC0037201      Control
10122.EC0044201     EC0044201 10122.EC0044201      Control
10122.EC0046201     EC0046201 10122.EC0046201      Control
10122.EC0051201     EC0051201 10122.EC0051201      Control


```{r setup,cache=TRUE}
#EXPLORE FOR OTU TABLE
otu_table(Dataset1)[1:5, 1:5]

# Total reads per sample
head(colSums(otu_table(Dataset1)))

# Total reads per taxon
head(rowSums(otu_table(Dataset1)))

# Dimensions
dim(otu_table(Dataset1))

```


```{r setup,cache=TRUE}
#EXPLORE

head(tax_table(Dataset1))


```



```{r setup,cache=TRUE}

library(ggplot2)


reads_df_1<-data.frame(
  reads=sample_sums(Dataset1),
  group=sample_data(Dataset1)$case_control
)

p1<-ggplot(reads_df, aes(x = reads, fill = group)) +
  geom_histogram(bins = 30, alpha = 0.7, color = "white") +
  scale_fill_manual(values = c("Case" = "#E74C3C", "Control" = "#3498DB")) +
  theme_minimal(base_size = 13) +
  labs(title = "Read Depth per Sample",
       x = "Total Reads", y = "Number of Samples", fill = "Group")

print(p)


#FOR DATASET 2 



reads_df_2<-data.frame(
  reads=sample_sums(Dataset2),
  group=sample_data(Dataset2)$case_control
)

p2<-ggplot(reads_df_2, aes(x = reads, fill = group)) +
  geom_histogram(bins = 30, alpha = 0.7, color = "white") +
  scale_fill_manual(values = c("Case" = "#E74C3C", "Control" = "#3498DB")) +
  theme_minimal(base_size = 13) +
  labs(title = "Read Depth per Sample",
       x = "Total Reads", y = "Number of Samples", fill = "Group")

#side by side
library(gridExtra)
merged<-grid.arrange(p1,p2,ncol=2)

print(merged)

ggsave("read_depth_both.png", plot = merged, width = 12, height = 5)





```



```{r setup ,cache=TRUE}
#Quantify the sparsity of the OTU table. so we need to convert the otu table to R expects matrix or #dataframe
otu1<-as(otu_table(Dataset1),"matrix")

otu2<-as(otu_table(Dataset2),"matrix")

zerosotu1<-sum(otu1==0)/length(otu1)

zerosotu2<-sum(otu2==0)/length(otu2)

#so 98.5% of Dataset 1's OTU table is zeros and 97.9% of Dataset 2's OTU table is zeros


#persample sparsity 
sample_sparsity_1<-colMeans(otu1==0) #colmean returns 0,1
sample_sparsity_2<-colMeans(otu2==0)
cat("\nDataset 1 - median fraction of ASVs absent per sample:",
    round(median(sample_sparsity_1), 3), "\n")
cat("Dataset 2 - median fraction of ASVs absent per sample:",
    round(median(sample_sparsity_2), 3), "\n")
```



we do not have enought metadata (for example age,sex,geographic location )
so will import the metadata from the original study and we will merge them based on the sample ID

```{r summary of phyloseq objects}
rank_names(Dataset1)     
sample_variables(Dataset1)

sra_meta<-read.csv("C:\\Users\\Νάσια\\Desktop\\volos summer school\\gut_microbiome_parkinson\\data\\SraRunTable.csv",check.names=FALSE,stringsAsFactor=FALSE)
cat("SRA metadata dimensions:", dim(sra_meta), "\n")
cat("Columns available:\n")
print(colnames(sra_meta))


```


```{r, explore the new metadata and merge}
#read csv
sra_meta<-read.csv("C:\\Users\\Νάσια\\Desktop\\volos summer school\\gut_microbiome_parkinson\\data\\SraRunTable.csv",check.names=FALSE,stringsAsFactor=FALSE)
cat("SRA metadata dimensions:", dim(sra_meta), "\n")
cat("Columns available:\n")
#print(colnames(sra_meta))

      
#split sra by dataset
sra_ds1<-sra_meta[sra_meta$Dataset=="dataset_1",]
sra_ds2<-sra_meta[sra_meta$Dataset=="dataset_2",]




#align and merge
#for dataset1 
idx1<-match(sample_names(Dataset1),sra_ds1$'Library Name') #find for each sample in phyloseq what row is in csv

meta1<-sra_ds1[idx1,]

rownames(meta1)<-sample_names(Dataset1)
sample_data(Dataset1)<-sample_data(meta1)

#for dataset2
idx2 <- match(sample_names(Dataset2), sra_ds2$`Library Name`)
meta2 <- sra_ds2[idx2, ]
rownames(meta2) <- sample_names(Dataset2)
sample_data(Dataset2) <- sample_data(meta2)


table(sample_data(Dataset1)$geographic_location)
table(sample_data(Dataset2)$geographic_location)


```



