---
title: "GUT MICROBIOME ML PROJECT parkinson"
output: html_document
date: "2026-04-12"
---

otu_table()Count matrixHow many of each bacteria per sample

sample_data()MetadataInfo ABOUT each patient

tax_table()TaxonomyWhat each bacteria is called

phy_tree()TreeEvolutionary relationships

```{r setup}




# ---- Libraries ----
library(phyloseq)

data_dir <- "Wallen_DAMethodCompare_2021/PhyloseqObjects"

# ---- Load phyloseq objects ----
library(here)
Dataset1 <- readRDS(here("data", "Dataset1", "phyloseq.rds"))
Dataset2 <- readRDS(here("data", "Dataset2", "phyloseq.rds"))


cat("Dataset1:", nsamples(Dataset1), "samples,", ntaxa(Dataset1), "taxa\n")
cat("Dataset2:", nsamples(Dataset2), "samples,", ntaxa(Dataset2), "taxa\n")
```



```{r setup,cache=TRUE}
# Step 1: Explore 
#FOR SAMPLE_DATA


table(sample_data(Dataset1)$case_control)

colnames(sample_data(Dataset1))

head(sample_data(Dataset1))


```
Sample Data:        [6 samples by 3 sample variables]:
                NGRC_SampleID        SampleID case_control
10122.EC0013201     EC0013201 10122.EC0013201      Control
10122.EC0037201     EC0037201 10122.EC0037201      Control
10122.EC0044201     EC0044201 10122.EC0044201      Control
10122.EC0046201     EC0046201 10122.EC0046201      Control
10122.EC0051201     EC0051201 10122.EC0051201      Control
```{r,finding NAs}

colSums(is.na(meta1))
colSums(is.na(meta2))


```

```{r setup,cache=TRUE}
#EXPLORE FOR OTU TABLE
otu_table(Dataset1)[1:5, 1:5]

# Total reads per sample
head(colSums(otu_table(Dataset1)))

# Total reads per taxon
head(rowSums(otu_table(Dataset1)))

# Dimensions
dim(otu_table(Dataset1))

```


```{r setup,cache=TRUE}
#EXPLORE

head(tax_table(Dataset1))


```



```{r setup,cache=TRUE}

library(ggplot2)


reads_df_1<-data.frame(
  reads=sample_sums(Dataset1),
  group=sample_data(Dataset1)$case_control
)

p1<-ggplot(reads_df_1, aes(x = reads, fill = group)) +
  geom_histogram(bins = 30, alpha = 0.7, color = "white") +
  scale_fill_manual(values = c("Case" = "#E74C3C", "Control" = "#3498DB")) +
  theme_minimal(base_size = 13) +
  labs(title = "Read Depth per Sample",
       x = "Total Reads", y = "Number of Samples", fill = "Group")

print(p1)


#FOR DATASET 2 



reads_df_2<-data.frame(
  reads=sample_sums(Dataset2),
  group=sample_data(Dataset2)$case_control
)

p2<-ggplot(reads_df_2, aes(x = reads, fill = group)) +
  geom_histogram(bins = 30, alpha = 0.7, color = "white") +
  scale_fill_manual(values = c("Case" = "#E74C3C", "Control" = "#3498DB")) +
  theme_minimal(base_size = 13) +
  labs(title = "Read Depth per Sample",
       x = "Total Reads", y = "Number of Samples", fill = "Group")

#side by side
library(gridExtra)
merged<-grid.arrange(p1,p2,ncol=2)

print(merged)

ggsave("read_depth_both.png", plot = merged, width = 12, height = 5)





```



```{r setup ,cache=TRUE}
#Quantify the sparsity of the OTU table. so we need to convert the otu table to R expects matrix or #dataframe
otu1<-as(otu_table(Dataset1),"matrix")

otu2<-as(otu_table(Dataset2),"matrix")

zerosotu1<-sum(otu1==0)/length(otu1)

zerosotu2<-sum(otu2==0)/length(otu2)

#so 98.5% of Dataset 1's OTU table is zeros and 97.9% of Dataset 2's OTU table is zeros


#persample sparsity 
sample_sparsity_1<-colMeans(otu1==0) #colmean returns 0,1
sample_sparsity_2<-colMeans(otu2==0)
cat("\nDataset 1 - median fraction of ASVs absent per sample:",
    round(median(sample_sparsity_1), 3), "\n")
cat("Dataset 2 - median fraction of ASVs absent per sample:",
    round(median(sample_sparsity_2), 3), "\n")
```



we do not have enought metadata (for example age,sex,geographic location )
so will import the metadata from the original study and we will merge them based on the sample ID

```{r summary of phyloseq objects}
rank_names(Dataset1)     
sample_variables(Dataset1)

sra_meta<-read.csv("C:\\Users\\Νάσια\\Desktop\\volos summer school\\gut_microbiome_parkinson\\data\\SraRunTable.csv",check.names=FALSE,stringsAsFactor=FALSE)
cat("SRA metadata dimensions:", dim(sra_meta), "\n")
cat("Columns available:\n")
print(colnames(sra_meta))


```


```{r, explore the new metadata and merge}
#read csv
sra_meta<-read.csv("C:\\Users\\Νάσια\\Desktop\\volos summer school\\gut_microbiome_parkinson\\data\\SraRunTable.csv",check.names=FALSE,stringsAsFactor=FALSE)
cat("SRA metadata dimensions:", dim(sra_meta), "\n")
cat("Columns available:\n")
#print(colnames(sra_meta))

      
#split sra by dataset
sra_ds1<-sra_meta[sra_meta$Dataset=="dataset_1",]
sra_ds2<-sra_meta[sra_meta$Dataset=="dataset_2",]


#align and merge
#for dataset1 
idx1<-match(sample_names(Dataset1),sra_ds1$'Library Name') #find for each sample in phyloseq what row is in csv

meta1<-sra_ds1[idx1,]

rownames(meta1)<-sample_names(Dataset1)
sample_data(Dataset1)<-sample_data(meta1)

#for dataset2
idx2 <- match(sample_names(Dataset2), sra_ds2$`Library Name`)
meta2 <- sra_ds2[idx2, ]
rownames(meta2) <- sample_names(Dataset2)
sample_data(Dataset2) <- sample_data(meta2)


table(sample_data(Dataset1)$geographic_location)
table(sample_data(Dataset2)$geographic_location)


```
```{r,NORMALISATION}

if (!requireNamespace("BiocManager", quietly = TRUE)) install.packages("BiocManager")
BiocManager::install("microbiome")

library(microbiome)
library(phyloseq)
library(ggplot2)

#CLR TRANFORMATION USING THE MICROBIOME LIBRARY
Dataset1_clr<-microbiome::transform(Dataset1,"clr")
Dataset2_clr<-microbiome::transform(Dataset2,"clr")





```

```{r}

dim(otu_table(Dataset1))


```


```{r}

print(otu_table(Dataset1))

```



```{r, prevelance of each taxon}
#we use the raw counts because its a binary question and in the normalized the zero counts become...not zero

#DATASET 1
prev_df_1 <- data.frame(
  Prevalence = colSums(otu_table(Dataset1) > 0) / nsamples(Dataset1),
  TotalAbundance = taxa_sums(Dataset1),
  Phylum = tax_table(Dataset1)[, "Phylum"]
)


nrow(prev_df_1)  
head(prev_df_1) 

#DATASET 2
prev_df_2<-data.frame(
  Prevalence=colSums(otu_table(Dataset2)>0)/nsamples(Dataset2),
  TotalAbundance=taxa_sums(Dataset2),
  Phylum=tax_table(Dataset2)[,"Phylum"]
)



```


```{r}
#prevalence by condition 
library(phyloseq)
library(ggplot2)
library(dplyr)
library(RColorBrewer)


############################################# DATASET1################################################################
# subset
ds1_case    <- subset_samples(Dataset1, case_control == "Case")
ds1_control <- subset_samples(Dataset1, case_control == "Control")

# prevalence per group
prev_case    <- colSums(otu_table(ds1_case) > 0)    / nsamples(ds1_case)
prev_control <- colSums(otu_table(ds1_control) > 0) / nsamples(ds1_control)

# fix tax_table length mismatch
phylum_vec <- as.vector(tax_table(Dataset1)[taxa_names(Dataset1), "Phylum"])

# build dataframe
prev_by_cond <- data.frame(
  Prev_Case    = as.numeric(prev_case),
  Prev_Control = as.numeric(prev_control),
  TotalAbundance = taxa_sums(Dataset1),
  Phylum       = phylum_vec
)

# colors
phyla_3 <- unique(prev_by_cond$Phylum)
palette_colors_3 <- colorRampPalette(brewer.pal(12, "Paired"))(length(phyla_3))
names(palette_colors_3) <- phyla_3

# plot Case vs Control prevalence per taxon
prevalence_plot_by_cond <- ggplot(prev_by_cond, aes(x = Prev_Control, y = Prev_Case, color = Phylum)) +
  geom_point(alpha = 0.6, size = 1.5) +
  scale_color_manual(values = palette_colors_3) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
  theme_minimal() +
  labs(title = "Prevalence by Condition — Dataset 1",
       subtitle = "Red line = equal prevalence in Case and Control",
       x = "Prevalence in Control",
       y = "Prevalence in Case")

print(prevalence_plot_by_cond)


############################################# DATASET2################################################################
ds2_case    <- subset_samples(Dataset2, case_control == "Case")
ds2_control <- subset_samples(Dataset2, case_control == "Control")

# prevalence per group
prev_case2   <- colSums(otu_table(ds2_case) > 0)    / nsamples(ds2_case)
prev_control2 <- colSums(otu_table(ds2_control) > 0) / nsamples(ds2_control)

# fix tax_table length mismatch
phylum_vec2 <- as.vector(tax_table(Dataset2)[taxa_names(Dataset2), "Phylum"])

# build dataframe
prev_by_cond2 <- data.frame(
  Prev_Case    = as.numeric(prev_case2),
  Prev_Control = as.numeric(prev_control2),
  TotalAbundance = taxa_sums(Dataset2),
  Phylum      = phylum_vec2
)

# colors
phyla_4 <- unique(prev_by_cond2$Phylum)
palette_colors_4 <- colorRampPalette(brewer.pal(12, "Paired"))(length(phyla_4))
names(palette_colors_4) <- phyla_4

# plot Case vs Control prevalence per taxon
prevalence_plot_by_cond2 <- ggplot(prev_by_cond2, aes(x = Prev_Control, y = Prev_Case, color = Phylum)) +
  geom_point(alpha = 0.6, size = 1.5) +
  scale_color_manual(values = palette_colors_4) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
  theme_minimal() +
  labs(title = "Prevalence by Condition — Dataset 2",
       subtitle = "Red line = equal prevalence in Case and Control",
       x = "Prevalence in Control",
       y = "Prevalence in Case")

print(prevalence_plot_by_cond2)

ggsave("plots/prevalence_by_cond_ds1.png", plot = prevalence_plot_by_cond,  width = 8, height = 6)
ggsave("plots/prevalence_by_cond_ds2.png", plot = prevalence_plot_by_cond2, width = 8, height = 6)
dir.create("plotsYES", showWarnings = FALSE)
```


```{r,fig.width=14,fig.height=7,making plots for prevelance}

#PLOT THE PREVALENCE
library(ggplot2)

#get phyla 
#FOR DATASET 1
phyla_1<-unique(prev_df_1$Phylum)

print(phyla_1)
n<-length(phyla_1)
n
#i will generate colors for each phyla
library(RColorBrewer)
palette_colors_1<-colorRampPalette(brewer.pal(12,"Paired"))(n)

names(palette_colors)<-phyla_1


#
prevalence_plot_1<-ggplot(prev_df_1, aes(x = TotalAbundance, y = Prevalence, color = Phylum)) +  
  geom_point(alpha = 0.6, size = 1.5) +
  scale_x_log10() +
  scale_color_manual(values=palette_colors_1)+
  geom_hline(yintercept = 0.1, linetype = "dashed", color = "red") +
  theme_minimal() +
  labs(title = "Prevalence vs Abundance — Dataset 1",
       subtitle = "Red line = 10% prevalence threshold",
       x = "Total Abundance (log10)",
       y = "Prevalence (fraction of samples)")                                


#FOR DATASET 2
phyla_2<-unique(prev_df_2$Phylum)

k<-length(phyla_2)
print(k)

palette_colors_2<-colorRampPalette(brewer.pal(12,"Paired"))(k)
                                   
prevalence_plot_2<-ggplot(prev_df_2,aes(x = TotalAbundance, y = Prevalence, color = Phylum)) +  
  geom_point(alpha = 0.6, size = 1.5) +
  scale_x_log10() +
  scale_color_manual(values=palette_colors_2)+
  geom_hline(yintercept = 0.1, linetype = "dashed", color = "red") +
  theme_minimal() +
  labs(title = "Prevalence vs Abundance — Dataset 2",
       subtitle = "Red line = 10% prevalence threshold",
       x = "Total Abundance (log10)",
       y = "Prevalence (fraction of samples)")                                

                                   
print(prevalence_plot_1)
print(prevalence_plot_2)


```

```{r, filtering the rare ones}


#filter dataset1 ->keep taxa present in >10% of samples

prevalence_threshold<-0.1

taxa_to_keep_1<-rownames(prev_df_1)[prev_df_1$Prevalence>=prevalence_threshold]
#prunes =removes taxa
Dataset1_filtered<-prune_taxa(taxa_to_keep_1,Dataset1)


#filter dataset1 
taxa_to_keep_2 <- rownames(prev_df_2)[prev_df_2$Prevalence >= prevalence_threshold]

Dataset2_filtered<-prune_taxa(taxa_to_keep_2,Dataset2)


#plot again 
#phylum_vec1 <- as.vector(tax_table(Dataset1_filtered)[taxa_names(Dataset1_filtered), "Phylum"])
#phylum_vec2 <- as.vector(tax_table(Dataset2_filtered)[taxa_names(Dataset2_filtered), "Phylum"])
# Recompute prevalence dataframes from filtered objects
prev_df_1_filtered <- data.frame(
  Prevalence     = colSums(otu_table(Dataset1_filtered) > 0) / nsamples(Dataset1_filtered),
  TotalAbundance = taxa_sums(Dataset1_filtered),
  Phylum         = as.vector(tax_table(Dataset1_filtered)[taxa_names(Dataset1_filtered), "Phylum"])
)

prev_df_2_filtered <- data.frame(
  Prevalence     = colSums(otu_table(Dataset2_filtered) > 0) / nsamples(Dataset2_filtered),
  TotalAbundance = taxa_sums(Dataset2_filtered),
  Phylum         = as.vector(tax_table(Dataset2_filtered)[taxa_names(Dataset2_filtered), "Phylum"])
)


phyla_1 <- unique(prev_df_1_filtered$Phylum)
palette_colors_1 <- colorRampPalette(brewer.pal(12, "Paired"))(length(phyla_1))
names(palette_colors_1) <- phyla_1

prevalence_plot_1 <- ggplot(prev_df_1_filtered, aes(x = TotalAbundance, y = Prevalence, color = Phylum)) +
  geom_point(alpha = 0.6, size = 1.5) +
  scale_x_log10() +
  scale_color_manual(values = palette_colors_1) +
  geom_hline(yintercept = 0.1, linetype = "dashed", color = "red") +
  theme_minimal() +
  labs(title = "Prevalence vs Abundance — Dataset 1 (filtered)",
       x = "Total Abundance (log10)", y = "Prevalence (fraction of samples)")

phyla_2 <- unique(prev_df_2_filtered$Phylum)
palette_colors_2 <- colorRampPalette(brewer.pal(12, "Paired"))(length(phyla_2))
names(palette_colors_2) <- phyla_2

prevalence_plot_2 <- ggplot(prev_df_2_filtered, aes(x = TotalAbundance, y = Prevalence, color = Phylum)) +
  geom_point(alpha = 0.6, size = 1.5) +
  scale_x_log10() +
  scale_color_manual(values = palette_colors_2) +
  geom_hline(yintercept = 0.1, linetype = "dashed", color = "red") +
  theme_minimal() +
  labs(title = "Prevalence vs Abundance — Dataset 2 (filtered)",
       x = "Total Abundance (log10)", y = "Prevalence (fraction of samples)")

prevalence_plot_1
prevalence_plot_2

```

```{r, CLR TRANSFORMATION}
#CLR(x) = log(x / geometric_mean(x))


#extract the OTU TABLE AGAIN as matrix (samples x taxa)
#as.matrix ->because we need to strip down only to numbers
Dataset1_clr<-microbiome::transform(Dataset1_filtered,"clr")
Dataset2_clr<-microbiome::transform(Dataset2_filtered,"clr")




```