---
title: "GUT MICROBIOME ML PROJECT parkinson"
output:
  pdf_document: default
  html_document: default
date: "2026-04-12"
---

otu_table()Count matrixHow many of each bacteria per sample

sample_data()MetadataInfo ABOUT each patient

tax_table()TaxonomyWhat each bacteria is called

phy_tree()TreeEvolutionary relationships

```{r setup}




# ---- Libraries ----
library(phyloseq)

data_dir <- "Wallen_DAMethodCompare_2021/PhyloseqObjects"

# ---- Load phyloseq objects ----
library(here)
Dataset1 <- readRDS(here("data", "Dataset1", "phyloseq.rds"))
Dataset2 <- readRDS(here("data", "Dataset2", "phyloseq.rds"))


cat("Dataset1:", nsamples(Dataset1), "samples,", ntaxa(Dataset1), "taxa\n")
cat("Dataset2:", nsamples(Dataset2), "samples,", ntaxa(Dataset2), "taxa\n")
```



```{r setup,cache=TRUE}
# Step 1: Explore 
#FOR SAMPLE_DATA


table(sample_data(Dataset1)$case_control)

colnames(sample_data(Dataset1))

head(sample_data(Dataset1))


```
Sample Data:        [6 samples by 3 sample variables]:
                NGRC_SampleID        SampleID case_control
10122.EC0013201     EC0013201 10122.EC0013201      Control
10122.EC0037201     EC0037201 10122.EC0037201      Control
10122.EC0044201     EC0044201 10122.EC0044201      Control
10122.EC0046201     EC0046201 10122.EC0046201      Control
10122.EC0051201     EC0051201 10122.EC0051201      Control
```{r,finding NAs}

#colSums(is.na(meta1))
#colSums(is.na(meta2))


```

```{r setup,cache=TRUE}
#EXPLORE FOR OTU TABLE
otu_table(Dataset1)[1:5, 1:5]

# Total reads per sample
head(colSums(otu_table(Dataset1)))

# Total reads per taxon
head(rowSums(otu_table(Dataset1)))

# Dimensions
dim(otu_table(Dataset1))

```


```{r setup,cache=TRUE}
#EXPLORE

head(tax_table(Dataset1))


```



```{r setup,cache=TRUE}

library(ggplot2)


reads_df_1<-data.frame(
  reads=sample_sums(Dataset1),
  group=sample_data(Dataset1)$case_control
)

p1<-ggplot(reads_df_1, aes(x = reads, fill = group)) +
  geom_histogram(bins = 30, alpha = 0.7, color = "white") +
  scale_fill_manual(values = c("Case" = "#E74C3C", "Control" = "#3498DB")) +
  theme_minimal(base_size = 13) +
  labs(title = "Read Depth per Sample",
       x = "Total Reads", y = "Number of Samples", fill = "Group")

print(p1)


#FOR DATASET 2 



reads_df_2<-data.frame(
  reads=sample_sums(Dataset2),
  group=sample_data(Dataset2)$case_control
)

p2<-ggplot(reads_df_2, aes(x = reads, fill = group)) +
  geom_histogram(bins = 30, alpha = 0.7, color = "white") +
  scale_fill_manual(values = c("Case" = "#E74C3C", "Control" = "#3498DB")) +
  theme_minimal(base_size = 13) +
  labs(title = "Read Depth per Sample",
       x = "Total Reads", y = "Number of Samples", fill = "Group")

#side by side
library(gridExtra)
merged<-grid.arrange(p1,p2,ncol=2)

print(merged)

ggsave("read_depth_both.png", plot = merged, width = 12, height = 5)





```



```{r setup ,cache=TRUE}
#Quantify the sparsity of the OTU table. so we need to convert the otu table to R expects matrix or #dataframe
otu1<-as(otu_table(Dataset1),"matrix")

otu2<-as(otu_table(Dataset2),"matrix")

zerosotu1<-sum(otu1==0)/length(otu1)

zerosotu2<-sum(otu2==0)/length(otu2)

#so 98.5% of Dataset 1's OTU table is zeros and 97.9% of Dataset 2's OTU table is zeros


#persample sparsity 
sample_sparsity_1<-colMeans(otu1==0) #colmean returns 0,1
sample_sparsity_2<-colMeans(otu2==0)
cat("\nDataset 1 - median fraction of ASVs absent per sample:",
    round(median(sample_sparsity_1), 3), "\n")
cat("Dataset 2 - median fraction of ASVs absent per sample:",
    round(median(sample_sparsity_2), 3), "\n")
```



we do not have enought metadata (for example age,sex,geographic location )
so will import the metadata from the original study and we will merge them based on the sample ID

```{r summary of phyloseq objects}
rank_names(Dataset1)     
sample_variables(Dataset1)

sra_meta<-read.csv("C:\\Users\\Νάσια\\Desktop\\volos summer school\\gut_microbiome_parkinson\\data\\SraRunTable.csv",check.names=FALSE,stringsAsFactor=FALSE)
cat("SRA metadata dimensions:", dim(sra_meta), "\n")
cat("Columns available:\n")
print(colnames(sra_meta))


```

```{r explore the new metadata and merge}
# read csv
sra_meta <- read.csv(
  "C:\\Users\\Νάσια\\Desktop\\volos summer school\\gut_microbiome_parkinson\\data\\SraRunTable.csv",
  check.names = FALSE, stringsAsFactors = FALSE
)
cat("SRA metadata dimensions:", dim(sra_meta), "\n")

# split sra by dataset
sra_ds1 <- sra_meta[sra_meta$Dataset == "dataset_1", ]
sra_ds2 <- sra_meta[sra_meta$Dataset == "dataset_2", ]

# align and merge
# for dataset1
idx1 <- match(sample_names(Dataset1), sra_ds1$`Library Name`)
meta1 <- sra_ds1[idx1, ]
rownames(meta1) <- sample_names(Dataset1)
sample_data(Dataset1) <- sample_data(meta1)

# for dataset2
idx2 <- match(sample_names(Dataset2), sra_ds2$`Library Name`)
meta2 <- sra_ds2[idx2, ]
rownames(meta2) <- sample_names(Dataset2)
sample_data(Dataset2) <- sample_data(meta2)

# ---- NEW: derive case_control from health_state ----
# SRA calls it "health_state" with values "Parkinsons"/"Control"
# Standardize to "case_control" with values "Case"/"Control"
sample_data(Dataset1)$case_control <- ifelse(
  sample_data(Dataset1)$health_state == "Parkinsons", "Case", "Control"
)
sample_data(Dataset2)$case_control <- ifelse(
  sample_data(Dataset2)$health_state == "Parkinsons", "Case", "Control"
)

# ---- VERIFY both variables exist and look right ----
cat("\n--- Dataset 1 ---\n")
print(table(sample_data(Dataset1)$case_control))
print(table(sample_data(Dataset1)$geographic_location))

cat("\n--- Dataset 2 ---\n")
print(table(sample_data(Dataset2)$case_control))
print(table(sample_data(Dataset2)$geographic_location))
```





```{r}

dim(otu_table(Dataset1))


```


```{r}

print(otu_table(Dataset1))

```



```{r, prevelance of each taxon}
#we use the raw counts because its a binary question and in the normalized the zero counts become...not zero

#DATASET 1
prev_df_1 <- data.frame(
  Prevalence = colSums(otu_table(Dataset1) > 0) / nsamples(Dataset1),
  TotalAbundance = taxa_sums(Dataset1),
  Phylum = tax_table(Dataset1)[, "Phylum"]
)


nrow(prev_df_1)  
head(prev_df_1) 

#DATASET 2
prev_df_2<-data.frame(
  Prevalence=colSums(otu_table(Dataset2)>0)/nsamples(Dataset2),
  TotalAbundance=taxa_sums(Dataset2),
  Phylum=tax_table(Dataset2)[,"Phylum"]
)



```


```{r}
#prevalence by condition 
library(phyloseq)
library(ggplot2)
library(dplyr)
library(RColorBrewer)


############################################# DATASET1################################################################
# subset
ds1_case    <- subset_samples(Dataset1, case_control == "Case")
ds1_control <- subset_samples(Dataset1, case_control == "Control")

# prevalence per group
prev_case    <- colSums(otu_table(ds1_case) > 0)    / nsamples(ds1_case)
prev_control <- colSums(otu_table(ds1_control) > 0) / nsamples(ds1_control)

# fix tax_table length mismatch
phylum_vec <- as.vector(tax_table(Dataset1)[taxa_names(Dataset1), "Phylum"])

# build dataframe
prev_by_cond <- data.frame(
  Prev_Case    = as.numeric(prev_case),
  Prev_Control = as.numeric(prev_control),
  TotalAbundance = taxa_sums(Dataset1),
  Phylum       = phylum_vec
)

# colors
phyla_3 <- unique(prev_by_cond$Phylum)
palette_colors_3 <- colorRampPalette(brewer.pal(12, "Paired"))(length(phyla_3))
names(palette_colors_3) <- phyla_3

# plot Case vs Control prevalence per taxon
prevalence_plot_by_cond <- ggplot(prev_by_cond, aes(x = Prev_Control, y = Prev_Case, color = Phylum)) +
  geom_point(alpha = 0.6, size = 1.5) +
  scale_color_manual(values = palette_colors_3) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
  theme_minimal() +
  labs(title = "Prevalence by Condition — Dataset 1",
       subtitle = "Red line = equal prevalence in Case and Control",
       x = "Prevalence in Control",
       y = "Prevalence in Case")

print(prevalence_plot_by_cond)

```

```{r ,prevalence by condition for dataset2}
############################################# DATASET2################################################################
ds2_case    <- subset_samples(Dataset2, case_control == "Case")
ds2_control <- subset_samples(Dataset2, case_control == "Control")

# prevalence per group
prev_case2   <- colSums(otu_table(ds2_case) > 0)    / nsamples(ds2_case)
prev_control2 <- colSums(otu_table(ds2_control) > 0) / nsamples(ds2_control)

# fix tax_table length mismatch
phylum_vec2 <- as.vector(tax_table(Dataset2)[taxa_names(Dataset2), "Phylum"])

# build dataframe
prev_by_cond2 <- data.frame(
  Prev_Case    = as.numeric(prev_case2),
  Prev_Control = as.numeric(prev_control2),
  TotalAbundance = taxa_sums(Dataset2),
  Phylum      = phylum_vec2
)

# colors
phyla_4 <- unique(prev_by_cond2$Phylum)
palette_colors_4 <- colorRampPalette(brewer.pal(12, "Paired"))(length(phyla_4))
names(palette_colors_4) <- phyla_4

# plot Case vs Control prevalence per taxon
prevalence_plot_by_cond2 <- ggplot(prev_by_cond2, aes(x = Prev_Control, y = Prev_Case, color = Phylum)) +
  geom_point(alpha = 0.6, size = 1.5) +
  scale_color_manual(values = palette_colors_4) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
  theme_minimal() +
  labs(title = "Prevalence by Condition — Dataset 2",
       subtitle = "Red line = equal prevalence in Case and Control",
       x = "Prevalence in Control",
       y = "Prevalence in Case")

print(prevalence_plot_by_cond2)

ggsave("plots/prevalence_by_cond_ds1.png", plot = prevalence_plot_by_cond,  width = 8, height = 6)
ggsave("plots/prevalence_by_cond_ds2.png", plot = prevalence_plot_by_cond2, width = 8, height = 6)
dir.create("plotsYES", showWarnings = FALSE)
```


```{r,fig.width=14,fig.height=7,making plots for prevelance}

#PLOT THE PREVALENCE
library(ggplot2)

#get phyla 
#FOR DATASET 1
phyla_1<-unique(prev_df_1$Phylum)

print(phyla_1)
n<-length(phyla_1)
n
#i will generate colors for each phyla
library(RColorBrewer)
palette_colors_1<-colorRampPalette(brewer.pal(12,"Paired"))(n)

names(palette_colors_1)<-phyla_1


#
prevalence_plot_1<-ggplot(prev_df_1, aes(x = TotalAbundance, y = Prevalence, color = Phylum)) +  
  geom_point(alpha = 0.6, size = 1.5) +
  scale_x_log10() +
  scale_color_manual(values=palette_colors_1)+
  geom_hline(yintercept = 0.1, linetype = "dashed", color = "red") +
  theme_minimal() +
  labs(title = "Prevalence vs Abundance — Dataset 1",
       subtitle = "Red line = 10% prevalence threshold",
       x = "Total Abundance (log10)",
       y = "Prevalence (fraction of samples)")                                


#FOR DATASET 2
phyla_2<-unique(prev_df_2$Phylum)

k<-length(phyla_2)
print(k)

palette_colors_2<-colorRampPalette(brewer.pal(12,"Paired"))(k)
                                   
prevalence_plot_2<-ggplot(prev_df_2,aes(x = TotalAbundance, y = Prevalence, color = Phylum)) +  
  geom_point(alpha = 0.6, size = 1.5) +
  scale_x_log10() +
  scale_color_manual(values=palette_colors_2)+
  geom_hline(yintercept = 0.1, linetype = "dashed", color = "red") +
  theme_minimal() +
  labs(title = "Prevalence vs Abundance — Dataset 2",
       subtitle = "Red line = 10% prevalence threshold",
       x = "Total Abundance (log10)",
       y = "Prevalence (fraction of samples)")                                

                                   
print(prevalence_plot_1)
print(prevalence_plot_2)


```

```{r, filtering the rare ones}


#filter dataset1 ->keep taxa present in >10% of samples

prevalence_threshold<-0.05

taxa_to_keep_1<-rownames(prev_df_1)[prev_df_1$Prevalence>=prevalence_threshold]
#prunes =removes taxa
Dataset1_filtered<-prune_taxa(taxa_to_keep_1,Dataset1)


#filter dataset1 
taxa_to_keep_2 <- rownames(prev_df_2)[prev_df_2$Prevalence >= prevalence_threshold]

Dataset2_filtered<-prune_taxa(taxa_to_keep_2,Dataset2)


#plot again 
#phylum_vec1 <- as.vector(tax_table(Dataset1_filtered)[taxa_names(Dataset1_filtered), "Phylum"])
#phylum_vec2 <- as.vector(tax_table(Dataset2_filtered)[taxa_names(Dataset2_filtered), "Phylum"])
# Recompute prevalence dataframes from filtered objects
prev_df_1_filtered <- data.frame(
  Prevalence     = colSums(otu_table(Dataset1_filtered) > 0) / nsamples(Dataset1_filtered),
  TotalAbundance = taxa_sums(Dataset1_filtered),
  Phylum         = as.vector(tax_table(Dataset1_filtered)[taxa_names(Dataset1_filtered), "Phylum"])
)

prev_df_2_filtered <- data.frame(
  Prevalence     = colSums(otu_table(Dataset2_filtered) > 0) / nsamples(Dataset2_filtered),
  TotalAbundance = taxa_sums(Dataset2_filtered),
  Phylum         = as.vector(tax_table(Dataset2_filtered)[taxa_names(Dataset2_filtered), "Phylum"])
)


phyla_1 <- unique(prev_df_1_filtered$Phylum)
palette_colors_1 <- colorRampPalette(brewer.pal(12, "Paired"))(length(phyla_1))
names(palette_colors_1) <- phyla_1

prevalence_plot_1 <- ggplot(prev_df_1_filtered, aes(x = TotalAbundance, y = Prevalence, color = Phylum)) +
  geom_point(alpha = 0.6, size = 1.5) +
  scale_x_log10() +
  scale_color_manual(values = palette_colors_1) +
  geom_hline(yintercept = 0.1, linetype = "dashed", color = "red") +
  theme_minimal() +
  labs(title = "Prevalence vs Abundance — Dataset 1 (filtered)",
       x = "Total Abundance (log10)", y = "Prevalence (fraction of samples)")

phyla_2 <- unique(prev_df_2_filtered$Phylum)
palette_colors_2 <- colorRampPalette(brewer.pal(12, "Paired"))(length(phyla_2))
names(palette_colors_2) <- phyla_2

prevalence_plot_2 <- ggplot(prev_df_2_filtered, aes(x = TotalAbundance, y = Prevalence, color = Phylum)) +
  geom_point(alpha = 0.6, size = 1.5) +
  scale_x_log10() +
  scale_color_manual(values = palette_colors_2) +
  geom_hline(yintercept = 0.1, linetype = "dashed", color = "red") +
  theme_minimal() +
  labs(title = "Prevalence vs Abundance — Dataset 2 (filtered)",
       x = "Total Abundance (log10)", y = "Prevalence (fraction of samples)")

prevalence_plot_1
prevalence_plot_2

```


```{r, ALPHA DIVERSITY}
# calculate alpha diversity on FILTERED datasets
alpha1 <- estimate_richness(Dataset1_filtered,
                            measures = c("Observed","Chao1","Shannon","Simpson","InvSimpson","Fisher"))
alpha2 <- estimate_richness(Dataset2_filtered,
                            measures = c("Observed","Chao1","Shannon","Simpson","InvSimpson","Fisher"))

# rownames to columns
alpha1$SampleID <- rownames(alpha1)
alpha1$Status <- sample_data(Dataset1_filtered)$case_control
alpha2$SampleID <- rownames(alpha2)
alpha2$Status <- sample_data(Dataset2_filtered)$case_control

library(ggplot2)
library(tidyr)

# combine both datasets for plotting
alpha1$Dataset <- "Dataset1"
alpha2$Dataset <- "Dataset2"
alpha_combined <- rbind(alpha1, alpha2)

# reshape to long format for faceting
alpha_long <- pivot_longer(alpha_combined,
                           cols = c("Observed","Shannon","Simpson"),
                           names_to = "Metric",
                           values_to = "Value")

# plot
ggplot(alpha_long, aes(x = Status, y = Value, fill = Status)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.7) +
  geom_jitter(width = 0.2, size = 0.8, alpha = 0.5) +
  facet_wrap(~ Metric + Dataset, scales = "free_y") +
  theme_minimal() +
  labs(title = "Alpha Diversity: PD vs Control")
#geom_jitter ->is like for the dots inside ,the outliers is outside and we dont want them so we put outliers.shape=NA
 #facet_wraps ->merged them ,splits by metric ,with scales to free_y because every metric has its own y axis 
                      




```
```{r,testing for differences in alpha diversity }
library(dplyr)

#For each Dataset + Metric combination, calculate the median for Cases, the median for #Controls, and the Wilcoxon p value comparing them."
#wilcox ->difference in medians
alpha_stas<-alpha_long %>%
  group_by(Dataset,Metric)%>%
  summarise(
    Case_median=median(Value[Status=="Case"],na.rm=TRUE),
    Control_median=median(Value[Status=="Control"],na.rm=TRUE),
    p_value=wilcox.test(Value[Status == "Case"], 
                                 Value[Status == "Control"])$p.value,
    .groups = "drop"
  )
  
#apply benjamini-honchberg FDR for multiple correction across al test

alpha_stas$p_adjusted<-p.adjust(alpha_stas$p_value,method="BH")


print(alpha_stas)



```
Alpha diversity shows no significant differences between PD and Control after FDR correction in either dataset, suggesting PD-associated microbiome changes are compositional rather than diversity-based, motivating beta diversity and differential abundance analysis.




```{r, CLR,aitchison,transform it to regular euclidean space where standard statistics actually work TRANSFORMATION}
#CLR(x) = log(x / geometric_mean(x))

#after a-diversity metrics 
#extract the OTU TABLE AGAIN as matrix (samples x taxa)

Dataset1_clr<-microbiome::transform(Dataset1_filtered,"clr")

Dataset2_clr<-microbiome::transform(Dataset2_filtered,"clr")



```




```{r, beta diversity,pcoa}
library(phyloseq)
library(ggplot2)
#bray curtis 
#it needs relative abundance not clr 

Dataset_1_rel<-transform_sample_counts(Dataset1_filtered,function(x) x/sum(x))
Dataset_2_rel<-transform_sample_counts(Dataset2_filtered,function(x) x/sum(x))

#relative abundance ->each sample must sum to 1
range(sample_sums(Dataset_1_rel))#range it returns the min and the max 
range(sample_sums(Dataset_2_rel))

#bray-curtis distance +pcoa 

ord1<-ordinate(Dataset_1_rel,method="PCoA",distance="bray")
ord2<-ordinate(Dataset_2_rel,method="PCoA",distance="bray")

#variance explained by each axis
vars_ds1<-round(ord1$values$Relative_eig[1:2]*100,1)
vars_ds2<-round(ord2$values$Relative_eig[1:2]*100,1)
cat("Dataset 1 — Axis1:", vars_ds1[1], "%, Axis2:", vars_ds1[2], "%\n")
cat("Dataset 2 — Axis1:", vars_ds2[1], "%, Axis2:", vars_ds2[2], "%\n")
#stat_ellipse=>It takes each color group (Case and Control) separately.
#For each group it fits a bivariate normal distribution to their (PCoA1, PCoA2) coordinates.
#It draws the 95% confidence ellipse — the region that should contain ~95% of points if the group were truly bivariate normal.

# ---- 4. Plot PCoA ----
p_beta1 <- plot_ordination(Dataset_1_rel, ord1, color = "case_control") +
  geom_point(size = 2, alpha = 0.7) +
  stat_ellipse(level = 0.95, linetype = "dashed") +
  scale_color_manual(values = c("Case" = "#E74C3C", "Control" = "#3498DB")) +
  theme_minimal(base_size = 13) +
  labs(title = "PCoA (Bray-Curtis) — Dataset 1",
       x = paste0("PCoA1 (", vars_ds1[1], "%)"),
       y = paste0("PCoA2 (", vars_ds1[2], "%)"),
       color = "Group")

p_beta2 <- plot_ordination(Dataset_2_rel, ord2, color = "case_control") +
  geom_point(size = 2, alpha = 0.7) +
  stat_ellipse(level = 0.95, linetype = "dashed") +
  scale_color_manual(values = c("Case" = "#E74C3C", "Control" = "#3498DB")) +
  theme_minimal(base_size = 13) +
  labs(title = "PCoA (Bray-Curtis) — Dataset 2",
       x = paste0("PCoA1 (", vars_ds2[1], "%)"),
       y = paste0("PCoA2 (", vars_ds2[2], "%)"),
       color = "Group")

# Side by side
print(p_beta1)
print(p_beta2)


```
```{r}

# extract the bray-curtis distance matrix already computed inside ord1/ord2
# no need to recompute - phyloseq stored it
bc_dist1<-phyloseq::distance(Dataset_1_rel,method="bray")
bc_dist2<-phyloseq::distance(Dataset_2_rel,method="bray")

#extract grouping variable
group1<-sample_data(Dataset_1_rel)$case_control
group2<-sample_data(Dataset_2_rel)$case_control
```



```{r, betadisper - check PERMANOVA assumption}
#betadisper creates an R object with multiple slots

#bd1$vectors → a matrix of PCoA coordinates (rows = samples, columns = PCoA axes)
#bd1$group → the vector of group labels (Case/Control) for each sample


# test whether group dispersions are equal
#Takes your Bray-Curtis distance matrix and the group labels (Case/Control). Calculates the distance of each sample from its group #centroid
bd1 <- betadisper(bc_dist1, group1)
bd2 <- betadisper(bc_dist2, group2)

print(bd1)
print(bd2)

#PLOTTING THE BETADISPER
#extract coordinates from betadisper object




#are the average dispersions significantly different between Case and Control? Uses permutations, same logic as PERMANOVA.
# p > 0.05 = dispersions are equal = PERMANOVA result is trustworthy
permutest(bd1)
permutest(bd2)

```
Average distance to median:

Case = 0.5719
Control = 0.5472

This is the actual dispersion measurement: the mean distance of each sample from its group's centroid in PCoA space. Cases are slightly more spread out than Controls (0.57 vs 0.55).
The difference looks tiny  only 0.025 but with 333 samples, even small differences become statistically detectable. That is why our permutest(bd1) gave p = 0.005. The biological effect is small, but real.



No. of Positive Eigenvalues: 177
No. of Negative Eigenvalues: 155

 Bray-Curtis is not a true Euclidean distance — it is semi-metric. When you do PCoA on it, you can get negative eigenvalues, which is  (negative variance does not exist in real space).
betadisper reports both counts so you know how "non-Euclidean" our distance matrix is. With 155/332 negative eigenvalues, our matrix has substantial non-Euclidean structure.

This is another reason Aitchison distance (=Euclidean on CLR) is methodologically cleaner. Aitchison gives you only positive eigenvalues — no negative weirdness. When you run PERMDISP on Aitchison distance later, you will see this difference.

FROM THE eigenvalues we see that non axis carries the signal




there is not separation 


```{r, PERMANOVA bray-curtis}
library(vegan)


#PERMANOVA---999 permutations is the standard
set.seed(42)
perm1<-adonis2(bc_dist1~group1,permutations = 999)
perm2<-adonis2(bc_dist2~group2,permutations = 999)
               
cat("=== Dataset 1 ===\n")
print(perm1)

cat("\n=== Dataset 2 ===\n")
print(perm2)
```
PERMANOVA has an assumption: the variance (spread) within each group must be similar. If Cases are much more spread out than Controls in your PCoA, PERMANOVA can give a significant p-value not because the groups are in different locations, but because one group is simply more dispersed. The test for this is called betadisper.

#adonis2 ITS partitioning the total variance in my distance matrix into
-how much is explained by group (case and control)
-how much is unexplained->the residual row

in the sumofsquares the residual is the variance not explained by p

R² = 0.00918 = 0.92%
This is the most important number. PD status explains 0.92% of total microbiome variance. Everything else — diet, age, geography, individual variation — explains the remaining 99.08%.

F = 3.07
The ratio of explained to unexplained variance, scaled by degrees of freedom.  Higher = stronger group separation relative to within-group scatter.


in dataset2 ->
R² = 0.57% → even smaller effect in Dataset 2
p = 0.003 → still significant
F = 2.88 → slightly weaker signal than Dataset 1

in the dataset2 we have more imbalanced groups that can affect F slightly 


PERMANOVA detected significant compositional differences between PD and Control samples in both datasets (Dataset 1: R² = 0.92%, p = 0.001; Dataset 2: R² = 0.57%, p = 0.003), with effect sizes consistent with the cross-study meta-analytic estimate of R² ≤ 1% reported by Romano et al. 2025. PERMDISP analysis revealed significant heterogeneity of dispersion in both datasets (Dataset 1: p = 0.005; Dataset 2: p = 0.001), indicating that PD samples exhibit greater compositional variability than Controls. This is biologically consistent with the heterogeneous nature of PD-associated dysbiosis, where disease state, medication, and gastrointestinal symptoms produce diverse microbial profiles. The small but reproducible PERMANOVA signal supports the existence of PD-associated compositional shifts, while the PERMDISP finding motivates supervised machine learning approaches that can leverage individual-level patterns rather than relying on group centroid separation

Comprehensive confounder adjustment as performed in the original Wallen et al. study was not possible due to limited public metadata availability 

```{r ,pca on aitchison distance}
#WE JUST USED PCOA WITH DIFFERENCE DISTANCE
#(Aitchison = Euclidean distance on CLR-transformed data)
library(vegan)
# Aitchison = Euclidean distance on CLR-transformed data
#  phyloseq stores taxa as rows by default → must transpose so rows = samples
# vegdist() computes distances between rows, so rows MUST be samples
euclidean_dist_1<-vegan::vegdist(t(otu_table(Dataset1_clr)),method="euclidean")


euclidean_dist_2<-vegan::vegdist(t(otu_table(Dataset2_clr)),method="euclidean")


#pca

# Aitchison distance computed in previous chunk: aitch_dist_1, aitch_dist_2

# PCoA on Aitchison distance (= PCA on CLR data, mathematically identical)
aitch_pcoa_1 <- ordinate(Dataset1_clr, method = "PCoA", distance = euclidean_dist_1)
aitch_pcoa_2 <- ordinate(Dataset2_clr, method = "PCoA", distance = euclidean_dist_2)




library(ggplot2)
library(patchwork)

# Variance explained by first 2 axes (for axis labels)
vars_aitc_1 <- round(aitch_pcoa_1$values$Relative_eig[1:2] * 100, 2)
vars_aitc_2 <- round(aitch_pcoa_2$values$Relative_eig[1:2] * 100, 2)

# Plot Dataset 1
p_aitc_beta1 <- plot_ordination(Dataset1_filtered, aitch_pcoa_1, color = "case_control") +
  geom_point(size = 2, alpha = 0.7) +
  stat_ellipse(level = 0.95, linetype = "dashed") +
  scale_color_manual(values = c("Case" = "#E74C3C", "Control" = "#3498DB")) +
  theme_minimal(base_size = 13) +
  labs(title = "PCoA (Aitchison) — Dataset 1",
       x = paste0("PCoA1 (", vars_aitc_1[1], "%)"),
       y = paste0("PCoA2 (", vars_aitc_1[2], "%)"),
       color = "Group")

# Plot Dataset 2
p_aitc_beta2 <- plot_ordination(Dataset2_filtered, aitch_pcoa_2, color = "case_control") +
  geom_point(size = 2, alpha = 0.7) +
  stat_ellipse(level = 0.95, linetype = "dashed") +
  scale_color_manual(values = c("Case" = "#E74C3C", "Control" = "#3498DB")) +
  theme_minimal(base_size = 13) +
  labs(title = "PCoA (Aitchison) — Dataset 2",
       x = paste0("PCoA1 (", vars_aitc_2[1], "%)"),
       y = paste0("PCoA2 (", vars_aitc_2[2], "%)"),
       color = "Group")

# Side by side using patchwork
p_aitc_beta1 + p_aitc_beta2


```



ANCOM-BC → takes raw counts (does its own internal transformation)
Wilcoxon on CLR → takes CLR values directly (it's just a rank test, doesn't care about sign)
DESeq2 → takes raw counts (negative binomial model, breaks on negatives)

```{r,DA}










